---
title: "Exercise 2"
author: "Holub Martin"
date: "25 9 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis

Do an exploratory data analysis of a matrix of expression values. Load the data and display:
* distributions: *boxplot*, *density*, *limma::plotDensities*
* normalization: *limma::normalizeQuantiles*
* clustering: *hclust*
* heatmap: *heatmap.2* or *pheatmap*
* correlation matrix: *cor* and *image*
* reduced dimensionality representation: *cmdscale* and *prcomp*


## Data Import
```{r }
anno = read.table("SampleAnnotation.txt", as.is=TRUE, sep="\t", quote="",
                  row.names=1, header=TRUE)
x = read.table("expressionData.txt", as.is=TRUE, sep="\t", quote="", row.names=1, header=TRUE, check.names = FALSE)
x = as.matrix(x)
```

## Define samples and colors and phenotype
```{r}
samples = rownames(anno)
colors = rainbow(nrow(anno))
isNorm = anno$TissueType == "norm"
isSick = anno$TissueType == "sick"
isAcute = anno$TissueType == "acute"
```
## Boxplot on a gene expression matrix

First transfrom with log2.

```{r}
boxplot(log2(x), use.cols = TRUE)
```

We see that the boxplot for all samples looks similarly, unsurprisingly as it is aggregate of `r length(unique(rownames(x)))` measurements. We look at othher exploratory plot.

## Kernel Density Estimation
Plot density estimates of log2 transformed gene expression levelsfor all samples
```{r}
for (i in 1:dim(x)[2]){
  df = density(log2(x[, i]), bw ="SJ", kernel= "gauss", na.rm = TRUE)
  color = colors[i]
  name = samples[i]
  if (i == 1){
    plot(df, col = color)
  } else {
    lines(df, col = color)
  } }
legend("topright", legend = samples, col = colors, lwd = 2.5, ncol = 2)
```

### Use limma::PlotDensities function
Here we can obtain the same plot with an one liner.
Correspondingly with the boxplot we see that mo stof the mass is located at lower value of expression levels (corresponding to 2-4 fold change). We also notice that there is a bump in the distribution around 8 fold change.
```{r}
limma::plotDensities(object = log2(x), col = colors, legend = "topright")
```
## Normalize data

Normalize columns of a matrix to have same quantiles and visualize densities. We see that they overlap for all samples indicating quantile allignement. We will keep working with the normalized data.

```{r}
x_norm <- limma::normalizeQuantiles((x))
limma::plotDensities(object = log2(x_norm), col = colors, legend = "topright")
```

## Hiearchical clustering

Pick a gene and look at partitioning to clusters. We observe that sick+acute and normal conditions appear together in respective clusters. Sick_14 appears to be an outlier. We use 1 - cor(x,y) as a distance metric, as discussed in lecture.

```{r}
# dist_measure <- dist(x_norm[1:10, ], method = "euclidian")
dist_measure <- as.dist((1-cor(x_norm))/2)
clusters <- hclust(dist_measure)
plot(clusters)
```

## Heatmap

Next we plot a heatmap

```{r}
dist_fun <-function(data){
  # Distance measure based on correlation
  return(as.dist((1-cor( t(data)) )/2))
}

gplots::heatmap.2(x_norm, distfun = dist_fun, col = colors, dendrogram = "row")

```